Do researchers preferentially collaborate with colleagues of the same gender?

### 2018-05-27 manuscript/Manuscript_Claire.Rmd
Some writing

---
title: "Do researchers preferentially collaborate with same-gendered colleagues?"
author:
- Luke Holman* and Claire Morandin&sect;
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage{microtype} 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Conference gender ratios} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and often feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping redress gender imbalance in the academic workforce, increasing women's access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this 'gender homophily' is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor, as predicted if mixed-gender teams produce better research. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

@@

\maketitle\unmarkedfntext{*School of BioSciences, The University of Melbourne, Victoria, Australia}
\maketitle\unmarkedfntext{§Organismal and Evolutionary Biology Research Programme,
Faculty of Biological and Environmental Sciences, University of Helsinki, Finland}
\newpage

# Introduction
Women are substantially underrepresented in many branches of the workforce in science, technology, engineering, mathematics, and medicine (STEMM), and face additional challenges and inequities relative to men [e.g. @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions [@Wutte_2007; @Reuben_2014] with lower salaries [@Trower_2002; @Umbach_2007], receive less grant money [@Hosek_2005; @OER_2008], are promoted more slowly [@Zuckerman_1987; @Rosenfeld_1991; @Long_1993; @Hopkins_2013], and are allocated fewer resources [@ODorchai_2009] and less research funding [@Feldt_1986, @Stack_2004, @Lariviere_2011] than men. Experimental studies have shown that people regard women's research achievements less favourably than identical achievements by men [@Moss_2012; @Knobloch_2013].

Because publishing, networking and collaboration are all instrumental to scientific productivity and thus, academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], dozens of studies have tested for gender differences in these areas [reviewed in Table S1 of @Holman_2018; see also @Bozeman_2011; @Abramo_2013; @Badar_2013; @Long_1992]. For example, studies have shown that relative to men, women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less with researchers from their own university department [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in general [@Fuchs_2001]. These gender differences in collaboration practice presumably have multiple concurrent causes, which might include implicit and explicit gender bias [REFS], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], concerns relating to sexual harassment [REFS], lower participation of women in international conferences [@Lewison_2001; Eisen 2014; @Martin_2014], and reduced receipt of research travel funds by women [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008; @Cohn_2014]. Additionally, empirical studies suggest that mixed-gender [@Bear_2011; @Campbell_2013] or otherwise diverse [@Hong_2004] teams can produce better results on collaborative tasks than less diverse teams. For reasons such as these, several studies have tested for gender differences in collaboration frequency or pattern by examining the author lists of published research. To our knowledge, every published study of this question has concluded that men co-publish with other men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015;@Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing has been termed 'gender homophily'.

However, we believe that these prior studies of gender homophily were hindered by a largely unacknowledged statistical issue that we term the Wahlund effect (\autoref{wahlund_plot}), meaning that the true extent of homophily is difficult to ascertain from current data. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if researchers actually prefer opposite-gendered colleagues. For example, a sample of literature containing bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018]. We chose the name by analogy with the Wahlund effect in population genetics [@Wahlund_1928], whereby spatial differences in allele frequencies create an excess of homozygotes across the whole population, even if alleles are randomly assorted within each subpopulation (original idea in BERGSTROM REF).

@@

# Methods
## Obtaining author lists and assigning gender
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every single article indexed on PubMed, from the oldest articles up to the present (which was August 2016), and attempting to assign a gender to each author based on their given name, using the web service genderize.io. Each journal was assigned to one of 107 scientific disciplines (e.g. _Nephrology_ or _Cell Biology_), using PubMed's journal categorisations where available. Because the present study focuses on co-authorship, all single-author papers were removed from the dataset. To simplify the statistical analysis, we also discarded all papers for which we could not determine the gender of every author with at least 95% certainty. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published either 0-1 or 10-11 years before the PubMed data were collected (i.e. $20^{th}$ August 2016). Lastly, we excluded journals with fewer than 50 suitable papers. This left us with a dataset from August 2015 - August 2016 containing `r sample.size$n2015[3]` papers with `r sample.size$n2015[4]` authorships, and a dataset from August 2005 - August 2006 containing `r sample.size$n2005[3]` papers and `r sample.size$n2005[4]` authorships. These papers came from `r sample.size$n2015[2]` and `r sample.size$n2005[2]` journals respectively, which were grouped into `r sample.size$n2015[1]` and `r sample.size$n2005[1]` research disciplines. There was a median of `r sample.size$n2015[5]` (`r sample.size$n2005[5]`) papers per journal, `r sample.size$n2015[6]` (`r sample.size$n2005[6]`) authors per journal, and `r sample.size$n2015[7]` (`r sample.size$n2005[7]`) authors per paper (the first number is for 2015-6, and the bracketed number for 2005-6).

@@

To estimate $\alpha$ for a particular subset of the scientific literature, we estimated $p$ as the average proportion of men's co-authors who are men (averaged across all papers with at least one man author), and $q$ as the average proportion of women's co-authors who are men (averaged across all papers with at least one woman author). To estimate the 95% confidence intervals on $\alpha$ for a given set of $n$ papers, we sampled $n$ papers with replacement 1000 times, estimated $\alpha$ on each sample, and recorded the 95% quantiles of the 1000 resulting estimates.

As well as calculating $\alpha$ for all authors, we calculated $\alpha$ for first or last authors only. $\alpha$ was again defined as $p - q$, but this time $p$ was estimated as the average proportion of male co-authors on papers with a male first (or last) author, and $q$ was estimated as the average proportion of male co-authors on papers with female first (or last) authors. We did not calculate $\alpha$ for other authorship positions (e.g. second or third authors) because this would necessitate reducing the dataset to papers with a sufficiently long authorship list, complicating comparison with other authorship positions that were calculated using a different sample.

Our test assumes that the expected value of $\alpha$ is zero if authors randomly assort, but for small datasets this assumption is not always true (as pointed out by Carl T. Bergstrom in a blog post, http://www.eigenfactor.org/gender/assortativity/note_to_eisen.rtf). To borrow Prof. Bergstrom's example, consider a small research specialty comprising just two men and two women researchers, who have together produced six two-author papers: one paper in each of the six possible two-author combinations. For these six papers, $\alpha = -\frac{1}{3}$, even though same- and opposite-gendered coauthors were selected in equal proportion to their frequency in the pool of possible collaborators. To control for the fact that the null expectation for $\alpha$ is not necessarily zero (particularly for very small datasets), we devised an adjusted version of the coefficient of homophily, which we term $\alpha'$. Every time we calculated $\alpha$ for a set of papers, we also determined the expected value of $\alpha$ under the null hypothesis that authors assort randomly with respect to gender. This was accomplished by randomly permuting authors across papers 1000 times, recalculating $\alpha$, and taking the median. We then calculated $\alpha'$ by subtracting the null expectation for $\alpha$ from the observed value. We also used the null-simulated $\alpha$ values to calculate a two-tailed p-value for the observed value of $\alpha$; the p-value was defined as the proportion of null simulations for which $|\alpha_{null}| > |\alpha_{obs}|$. We applied false discovery rate correction to each set of p-values to account for multiple testing following Benjamini and Hochberg [-@Benjamini_1995].

As expected, $\alpha'$ was usually almost identical to $\alpha$ (Figure S1), but $\alpha$ was downwardly biased relative to $\alpha'$ for small datasets (Figure S2). Additionally, the correlation between $\alpha'$ and sample size was negligible ($R^2 < 0.01$), suggesting that our calculation of $\alpha'$ effectively removed the dependence of $\alpha$ on sample size. We therefore use the adjusted statistic $\alpha'$ in all analyses.

## Minimising the Wahlund effect: discipline and time period
As discussed in the Introduction and shown in \autoref{wahlund_plot}, the Wahlund effect can give the appearance of gender homophily (reflected in our study as $\alpha' > 0$) whenever one lumps together different 'subpopulations' of academic papers that differ in their author gender ratios.

To minimise bias in $\alpha'$ due to the Wahlund effect, we restricted each set of papers to a single research specialty to the greatest extent allowed by our data. Specifically, we only calculated $\alpha'$ for individual journals, since papers from the same journal typically focus on one or a few closely related topics. Although some journals, e.g. _PLoS ONE_, publish research from diverse disciplines with very different author gender ratios [@Holman_2018], calculating $\alpha'$ for these highly multidisciplinary journals is still useful as a contrast. The difference in $\alpha'$ between highly multidisciplinary and more specialised journals, e.g. _PLoS ONE_ versus _PLoS Computational Biology_, gives a rough estimate of the extent to which multidisciplinarity can inflate $\alpha'$.

As well as varying between disciplines, the gender ratio of authors has changed markedly over time [@Holman_2018]. Because the gender ratio was more male-biased in the past, $\alpha'$ would be inflated if we calculated it for a sample of papers published over a long time frame. To minimise this effect, we only sampled papers from two one-year periods (namely 2005-6 and 2015-16). The median change per year in % (fe)male authors across journals is < 0.5% [@Holman_2018], and so restricting our dataset to a single year should prevent temporal changes in gender ratio from noticeably inflating our estimates of $\alpha'$.

@@

## Calculating standardised journal impact factor
We obtained the 3-year impact factor for each journal from Clarivate Analytics. To account for large differences in impact factor between disciplines, we took the the residuals from a model with $Log_{10}$ impact factor as the response and the research discipline of the journal as a random effect. Thus, journals with a positive standardised impact factor have a higher mean number of citations than the average for journals in their discipline. We then used Spearman rank correlation to test whether our estimates of gender homophily covaried with journal impact factor.

## Statistical analysis
Previous authors (e.g. @Bonham_2017) have hypothesised that senior scientists preferentially recruit staff and students of the same gender, and/or that junior researchers preferentially select same-gendered mentors. In the majority of disciplines, authorship conventions mean that the first-listed author is often an early-career researcher, while the author listed last is more likely to be a senior researcher leading a research team [@Wren_2007]. Assuming that senior researchers are the main drivers of homophily and that there are enough papers with three or more authors, we predict that the last author's gender will be the strongest predictor of the remaining authors' genders (i.e. the gender of the last author will be more salient than that of the first author, or any other authorship position). This is because the first author's gender would simply be an imperfect correlate of the true causal effect, while the last author's gender would be the causal effect itself.

@@

## Theoretical expectations for $\alpha$ when there is a gender gap between career stages
In many STEMM subjects, the gender ratio is more skewed among established researchers relative to early-career researchers [@Shaw_2012; @Holman_2018]. We hypothesised that this skew could potentially create Wahlund effects and 'reverse' Wahlund effects. For example, imagine that the majority of collaborations are between students and professors, and that the gender ratio is skewed in opposite directions at each career stage: we will then see an excess of mixed-gender coauthorships (heterophily, $\alpha < 0$), even if gender has no causal effect on the selection of coauthors. Similarly, a hypothetical field in which students work only with students, and professors with professors, will have apparent gender homophily ($\alpha > 0$).

Because we can think of no tractable way to control for this issue using our dataset (which contains no information on career stage), we instead decided to derive the theoretical expectations for $\alpha$ when there is a difference in gender ratio across career stages. For simplicity, we assume there are only two career stages, though the results are likely to be general. Under the null model that gender has no causal effect on collaboration, we calculated $\alpha$ for various combinations of the four free parameters: the gender ratios for early- and late-career researchers, and the relative frequency of collaborations between early-early, early-late, and late-late collaborations. We then used the theoretical expectations for $\alpha$ to properly qualify our main conclusions (see Results). The R code used to derive the theoretical expectations can be found in the Online Supplementary Material.

## Data availability and reproducibility
The Online Supplementary Material is an HTML document prepared with R Markdown, which contains R scripts used to produce all data, figures and tables. The input data from Holman et al. [-@Holman_2018] is archived at https://osf.io/bt9ya/.

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. The great majority of journals had $\alpha' > 0$, and for many of these the corrected two-tailed p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6. The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, such that we could not reject the null hypothesis of random assortment with respect to gender.

@@

$\alpha'$ was significantly higher in the sample from 2015-16 relative to 2005-6, but the difference in means was small (Figure S3; Effect of the fixed factor 'Time period' in a linear mixed model of the data for all author positions: Cohen's $d$ = $0.09{\pm}0.04$, $t_{948}$ = 2.51, p = 0.012).

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (Figure S4; Effect of the fixed factor 'Authorship position' in a linear mixed model: Cohen's $d$ = $0.07{\pm}0.02$, $t_{1988}$ = 4.48, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors' genders than the gender of the last author, i.e. the opposite of our prediction based on the hypothesis that senior scientists are causally responsible for homophily.

## Variance in homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was $0.12{\pm}0.02$ for Urology journals, and $0.03{\pm}0.01$ for Veterinary Medicine journals; \autoref{alpha_histograms}, Table S3). However, there was little evidence for strong differences in $\alpha'$ between disciplines: the random factor 'Discipline' explained at most 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). This implies that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

@@

## Relationship between gender homophily and gender ratio
We next tested whether researchers are more or less likely to seek out same-gendered colleagues in strongly gender-biased disciplines (e.g. Surgery or Nursing), relative to disciplines with a comparatively gender-balanced workforce (e.g. Psychiatry). We found a positive, non-linear relationship between the overall gender ratio of all authors publishing in a particular journal [as estimated by @Holman_2018], and the estimated value of $\alpha'$ for all authors and for first authors (\autoref{alpha_gender_ratio}). Journals with a balanced or female-biased author gender ratio tended to have higher $\alpha'$ than journals with a male-biased author gender ratio (GAM smooth terms p < 0.001, see Online Supplementary Material). The relationship was not statistically significant when $\alpha'$ was calculated for last authors (GAM, p = 0.142), though the trend appeared similar (\autoref{alpha_gender_ratio}).

>>>>

---
title: "Do researchers preferentially collaborate with colleagues of the same gender?"
author:
- Luke Holman* and Claire Morandin&sect;
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage{microtype} 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Gender and coauthorship} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and often feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping to redress gender imbalance in the academic workforce, increasing women's access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this 'gender homophily' is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor, as predicted if mixed-gender teams produce better research. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

@@

\maketitle\unmarkedfntext{*School of BioSciences, The University of Melbourne, Victoria, Australia.}
\maketitle\unmarkedfntext{§Organismal and Evolutionary Biology Research Programme,
Faculty of Biological and Environmental Sciences, University of Helsinki, Finland.}

\newpage
# Introduction

- Remember dataset S1 issue (put in ESM?)
- re-run model on line 912 and 931 (weights issue)
- check figure legend line 900. Also check the 'significant' area hasn't changed loads
- proportional alpha?

Women are substantially underrepresented in many branches of the workforce in science, technology, engineering, mathematics, and medicine (STEMM), and face additional challenges and inequities relative to men [e.g. @Shaw_2012; @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions [@Wutte_2007; @Reuben_2014] with lower salaries [@Trower_2002; @Umbach_2007], receive less grant money [@Hosek_2005; @OER_2008], are promoted more slowly [@Zuckerman_1987; @Rosenfeld_1991; @Long_1993; @Hopkins_2013], and are allocated fewer resources [@ODorchai_2009] and less research funding [@Feldt_1986; @Stack_2004; @Lariviere_2011] than men. Experimental studies suggest that women's research achievements are regarded less favourably than identical achievements by men [@Moss_2012; @Knobloch_2013].

Because publishing, networking and collaboration are instrumental to scientific productivity and academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], dozens of studies have tested for gender differences in these areas [e.g. @Long_1992; @Bozeman_2011; @Abramo_2013; @Badar_2013; references in Table S1 of @Holman_2018]. For example, studies have found that women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less within their own university departments [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in total [@Fuchs_2001]. These gender differences in collaboration practice presumably have multiple concurrent causes, which might include implicit and explicit gender bias [@Moss_2012], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], concerns relating to sexual harassment [@jagsi_2016], and unequal access to conferences [@Martin_2014] and travel funds [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008; @Cohn_2014]. Additionally, empirical studies suggest that mixed-gender [@Bear_2011; @Campbell_2013] or otherwise diverse [@Hong_2004] teams can produce better results on collaborative tasks than less diverse teams. For reasons such as these, several studies have tested for gender differences in collaboration frequency or pattern by examining the author lists of published research. To our knowledge, every published study of this question suggests that men co-publish with other men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing has been termed 'gender homophily'.

However, we believe that prior studies of gender homophily were hindered by a largely unacknowledged statistical issue that we term the Wahlund effect (\autoref{wahlund_plot}), by analogy with a conceptually similar effect in population genetics [@Wahlund_1928]. The Wahlund effect makes it deceptively hard to measure gender homophily. Essentially, whenever coauthorship data are sampled from two or more discrete sets of literature, which vary in the author gender ratio and which are largely not connected by collaboration, the number of same-gendered coauthors will be inflated. This can give the impression that authors preferentially publish with same-gendered colleagues if no gender preferences exist, or even if researchers actually prefer opposite-gendered colleagues. For example, a sample of literature containing bioinformatics and cell biology papers will probably contain an excess of mostly-male and mostly-female author lists, simply because researchers preferentially work with colleagues from the same discipline, and because the author gender ratio is more male-biased in bioinformatics than in cell biology [@Holman_2018].

@@

# Methods
## The dataset
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every article indexed on PubMed, from the oldest articles up to the present, and attempting to assign a gender to each author based on their given name. Each journal was assigned to one of 107 scientific disciplines, using PubMed's journal categorisations where available. Because the present study focuses on co-authorship, all single-author papers were discarded. We also discarded all papers for which we could not determine the gender of every author with ${\ge}95\%$ certainty, in order to simplify the statistical analysis. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published either 0-1 or 10-11 years before the PubMed data were collected (i.e. 20^th^ August 2016). Lastly, we excluded journals with fewer than 50 suitable papers.

The final sample sizes were `r sample.size$n2015[3]` papers and `r sample.size$n2015[4]` authorships from August 2015 - August 2016, and `r sample.size$n2005[3]` papers and `r sample.size$n2005[4]` authorships from August 2005 - August 2006. Respectively, these papers came from `r sample.size$n2015[2]` and `r sample.size$n2005[2]` journals, which were grouped into `r sample.size$n2015[1]` and `r sample.size$n2005[1]` research disciplines. There was a median of `r sample.size$n2015[5]` (`r sample.size$n2005[5]`) papers per journal, `r sample.size$n2015[6]` (`r sample.size$n2005[6]`) authors per journal, and `r sample.size$n2015[7]` (`r sample.size$n2005[7]`) authors per paper (the first number is for 2015-6, and the bracketed number for 2005-6).

@@

To estimate $\alpha$ for a particular subset of the scientific literature, we estimated $p$ as the average proportion of men's co-authors who are men (averaged across all papers with at least one man author), and $q$ as the average proportion of women's co-authors who are men (averaged across all papers with at least one woman author). To estimate the 95% confidence intervals on $\alpha$ for a given set of $n$ papers, we sampled $n$ papers with replacement 1000 times, estimated $\alpha$ on each sample, and recorded the 95% quantiles of the resulting 1000 estimates.

As well as calculating $\alpha$ for all authors, we calculated $\alpha$ for first or last authors only. $\alpha$ was again defined as $p - q$, but this time $p$ was estimated as the average proportion of male co-authors on papers with a male first (or last) author, and $q$ was estimated as the average proportion of male co-authors on papers with female first (or last) authors. We did not calculate $\alpha$ for other authorship positions (e.g. second or third authors) because this would necessitate culling the dataset to include only papers with a sufficiently long author list, complicating interpretation of the results.

Our test assumes that the expected value of $\alpha$ is zero if authors randomly assort, but for small datasets this assumption is not always true (as pointed out by Carl T. Bergstrom in a blog post, http://www.eigenfactor.org/gender/assortativity/note_to_eisen.rtf). To borrow Prof. Bergstrom's example, consider a small research specialty comprising just two men and two women researchers, who have together produced six two-author papers: one in each of the six possible two-author combinations. For these six papers, $\alpha = -\frac{1}{3}$, even though same- and opposite-gendered coauthors were selected in equal proportion to their frequency in the pool of possible collaborators. To control for the fact that the null expectation for $\alpha$ is not zero for small datasets, we devised an adjusted version of the coefficient of homophily, which we term $\alpha'$. Every time we calculated $\alpha$ for a set of papers, we also determined the expected value of $\alpha$ under the null hypothesis that authors assort randomly with respect to gender. This was accomplished by randomly permuting authors across papers 1000 times, recalculating $\alpha$, and taking the median. We then calculated $\alpha'$ by subtracting the null expectation for $\alpha$ from the observed value. We also used the null-simulated $\alpha$ values to calculate a two-tailed p-value for the observed value of $\alpha$; the p-value was defined as the proportion of null simulations for which $|\alpha_{null}| > |\alpha_{obs}|$. We applied false discovery rate (FDR) correction to each set of p-values to account for multiple testing [@Benjamini_1995].

As expected, $\alpha'$ was usually almost identical to $\alpha$ (Figure S1), but $\alpha$ was downwardly biased relative to $\alpha'$ for small datasets (Figure S2). Additionally, the correlation between $\alpha'$ and sample size was negligible ($R^2 < 0.01$), suggesting that our calculation of $\alpha'$ effectively removed the dependence of $\alpha$ on sample size. We therefore used $\alpha'$ in all analyses.

## Minimising the Wahlund effect: research discipline and time period
To minimise bias in $\alpha'$ due to the Wahlund effect, we restricted each set of papers to a single research specialty to the greatest extent allowed by our data. Specifically, we only calculated $\alpha'$ for individual journals, since papers from the same journal typically focus on one or a few closely related topics. Although some journals, e.g. _PLoS ONE_, publish research from diverse disciplines with very different author gender ratios [@Holman_2018], calculating $\alpha'$ for these highly multidisciplinary journals is still useful as a contrast. The difference in $\alpha'$ between highly multidisciplinary and more specialised journals, e.g. _PLoS ONE_ versus _PLoS Computational Biology_, gives a rough estimate of the extent to which multidisciplinarity inflates $\alpha'$.

As well as varying between disciplines, the gender ratio of authors has changed markedly over time [@Holman_2018]. Because the gender ratio was more male-biased in the past, $\alpha'$ would be inflated if we calculated it for a sample of papers published over a long enough time frame. To minimise this effect, we only sampled papers from two one-year periods (namely 2005-6 and 2015-16). The median change per year in % (fe)male authors across journals is below 0.5% [@Holman_2018], and so restricting our dataset to a single year should prevent temporal changes in gender ratio from noticeably affecting our estimates of $\alpha'$.

@@

## Calculating standardised journal impact factor
We obtained the 3-year impact factor for each journal from Clarivate Analytics. To account for large differences in impact factor between disciplines, we took the the residuals from a model with $Log_{10}$ impact factor as the response and the research discipline of the journal as a random effect. Thus, journals with a positive standardised impact factor have a higher mean number of citations than the average for journals in their discipline. We then used Spearman rank correlation to test whether $\alpha'$ was correlated with impact factor across journals.

## Statistical analysis
Previous authors [e.g. @Bonham_2017] have hypothesised that senior scientists preferentially recruit staff and students of the same gender, and/or that junior researchers preferentially select same-gendered mentors. In the majority of disciplines, authorship conventions mean that the first-listed author is often an early-career researcher, while the author listed last is more likely to be a senior researcher leading a research team [@Wren_2007]. Assuming that senior researchers are the main drivers of homophily and that there are enough papers with three or more authors, we predict that the last author's gender will be the strongest predictor of the remaining authors' genders (i.e. the gender of the last author will be more salient than that of the first author, or any other authorship position). This is because the first author's gender would simply be an imperfect correlate of the true causal effect, while the last author's gender would be the causal effect itself.

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
In many STEMM subjects, the gender ratio is more skewed among established researchers relative to early-career researchers [@Shaw_2012; @Holman_2018]. We hypothesised that this skew could potentially create Wahlund effects and 'reverse' Wahlund effects. For example, imagine that the majority of collaborations are between students and professors, and that the gender ratio differs between career stages: we will then see an excess of mixed-gender coauthorships (heterophily, $\alpha < 0$), even if gender has no causal effect on the selection of coauthors. Similarly, a hypothetical field in which students work only with students, and professors with professors, would have apparent gender homophily ($\alpha > 0$).

Because we can think of no tractable method of controlling for this issue using our dataset, which contains no information on career stage. Therefore, we instead decided to derive the theoretical expectations for $\alpha$ when there is a difference in gender ratio across career stages. For simplicity, our calculations assume there are only two career stages, though the conclusions are likely to generalise to a more realistic multi-tiered career ladder. Under the null model that gender has no causal effect on collaboration, we calculated $\alpha$ for various combinations of the four free parameters, i.e. the gender ratios for early- and late-career researchers, and the relative frequency of collaborations between early-early, early-late, and late-late collaborations. We then used the theoretical expectations for $\alpha$ to qualify our main conclusions (see Results). Annotated R code used to derive the theoretical expectations can be found in the Online Supplementary Material.

## Data availability and reproducibility
The Online Supplementary Material contains R scripts used to produce all data, figures and tables. The input data from Holman et al. [-@Holman_2018] is archived at https://osf.io/bt9ya/.

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. The great majority of journals had $\alpha' > 0$, and for many of these the FDR-corrected p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6. The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, consistent with the null hypothesis of random assortment with respect to gender.

@@

$\alpha'$ was significantly higher in the literature sample from 2015-16 relative to 2005-6, but the difference in means was small (Figure S3; Effect of the fixed factor 'Time period' in a linear mixed model of the data for all author positions: Cohen's $d$ = $0.09{\pm}0.04$, $t_{948}$ = 2.51, p = 0.012).

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (Figure S4; Effect of the fixed factor 'Authorship position' in a linear mixed model: Cohen's $d$ = $0.07{\pm}0.02$, $t_{1988}$ = 4.48, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors' genders than the gender of the last author, i.e. the opposite of what is predicted if senior scientists are causally responsible for homophily.

## Variance in homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was $0.12{\pm}0.02$ for Urology journals, and $0.03{\pm}0.01$ for Veterinary Medicine journals; \autoref{alpha_histograms}, Table S3). However, there was no evidence for consistent differences in $\alpha'$ between disciplines: the random factor 'Discipline' explained at most 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). This implies that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

@@

## Relationship between gender homophily and gender ratio
We next tested whether researchers are more or less likely to publish with same-gendered colleagues in strongly gender-biased disciplines (e.g. Surgery or Nursing), relative to disciplines with a comparatively gender-balanced workforce (e.g. Psychiatry). We found a positive, non-linear relationship between the overall gender ratio of all authors publishing in a particular journal [as estimated in @Holman_2018], and the estimated value of $\alpha'$ for all authors and for first authors (\autoref{alpha_gender_ratio}). Journals with a balanced or female-biased author gender ratio tended to have higher $\alpha'$ than journals with a male-biased author gender ratio (GAM smooth terms p < 0.001; Online Supplementary Material). The relationship was not statistically significant when $\alpha'$ was calculated for last authors (GAM, p = 0.142), though the trend appeared similar (\autoref{alpha_gender_ratio}).
